Numerical simulation of the Cordilleran ice sheet through the last glacial cycle

### 2014-06-18 cordillera-cycle.tex
Polish deglaciation history.

In the North American Cordillera alike other glaciated regions, the large
majority of the glacial geologic record relates to the last few millennia of
glacier cover, most of the older evidence having been overridden by ice retreat
retreat during the deglaciation \footnote{\textbf{Johan,} can we refer to one
of your papers here?}.

@@

The possibly of a widespread late-glacial re-advance of decaying glaciers of
the North American Cordilleran has been debated for long. On the southern tip
of the Coast Mountains, radiocarbon-dated frontal moraines indicate multiple
stages of glacial re-advance in the Fraser and Squamish valleys, of which
one corresponds to the Younger Dryas chronozone \citep{Clague.etal.1997,
Friele.Clague.2002, Friele.Clague.2002a, Kovanen.2002,
Kovanen.Easterbrook.2002}. Additional evidence for a late-glacial re-advance
was found in the Finlay River area, northern Rocky Mountains, where
sharp-crested moraines show an abundant remobilization of late-glacial
sediments by alpine glaciers in interaction with the main body of the decaying
ice sheet \citep{Lakeman.etal.2008}. Although further work is needed to
constrain the timing of the North American Cordillera late-glacial re-advance,
to seize its extent and identify potential climatic triggers
\citep{Menounos.etal.2008}, it is interesting to note that the simulation
driven by the GRIP record produces a late-glacial re-advance in the Coast
Mountains, Rocky Mountains and Finlay River area, corresponding to where it
has been identified in the geologic record (Fig.~\ref{fig:deglac}, left panel),
whereas the EPICA-driven simulation produces a nearly-continuous deglaciation
with very restrained glacial re-advance (Fig.~\ref{fig:deglac}, right panel).

Although the predicted timing of glacier retreat differs between the two
simulations, its patterns are relatively consistent (Fig.~\ref{fig:deglac}). In
both runs, the southern half of the modelling domain, including the Puget
Lowland, the Coast and Rocky Mountains and the Interior Plateau of British
Columbia, are completely deglaciated by 12\,kyr, whereas significant ice cover
remains on the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias Mountains
in the northern half of the domain. After 12\,kyr, deglaciation proceeds
through an opening of the Liard lowland with radial margin retreat towards the
surrounding mountain ranges and inward flow towards the depression, in full
compatibility with the regional melt water record \citep{Margold.etal.2013}.
Remaining ice continues its decay by retreating towards the Selwyn Mountains
and, then, the Skeena Mountains. The last remains of the palaeo-ice sheet
vanish in the Skeena Mountains at around 10.6\,kyr in the EPICA simulation
versus 9.4\,kyr in the GRIP simulation.

Once more, snapshots of modelled deglacial ice sheet configurations
(Figs.~\ref{fig:deglacshots-grip} and~\ref{fig:deglacshots-epica}) provide a
poor medium for direct comparison with geomorphological evidence of the ice
retreat. Here, we compute a map of de-glacial, basal flow direction immediately
preceding exposure of ice-free land or cessation of sliding upon the approach
of the retreating margin (Fig.~\ref{fig:lastflow}). On the northern and
southern front of the ice sheet and in the Liard Lowland, the resulting map
show similarities with patterns of glacial lineations
(\citealp{Prest.etal.1968}; \citealp[Fig.~1.12]{Clague.1989};
\citealp[Fig.~2]{Kleman.etal.2010}; \citep[Fig.~2]{Margold.etal.2013}).
However, such is not the case on the Interior Plateau of British Columbia,
where the model produce no deglacial flow directions (Fig.~\ref{fig:lastflow}),
whereas glacial lineation maps indicate eastward flow.

On the one hand, this could indicate that the model performs poorly in this
region, perhaps because the modelled last glacial maximum ice sheet is too
thick, or too extensive to the west. Because feedback mechanisms between ice
sheet topography and regional climate such as wind redistribution, orographic
precipitation and latent warming of moisture-depleted air are absent in our
model, we regard this possibility as fully probable \citep{Seguinot.etal.2013}.

On the other hand, in both simulations driven by the GRIP and EPICA records,
the deglaciation of central British Columbia consists of a rapid northwards
retreat associated with southwards-flowing, non-sliding ice lobes on the
Interior Plateau, while the surrounding mountain ranges are already ice-free
(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). Thus the model
suggests that glacial lineations on the Interior Plateau of British Columbia
may be of older age than the last glacial maximum, and may have remained intact
throughout the deglaciation.

>>>>

In the North American Cordillera alike other glaciated regions, the large
majority of the glacial geologic record relates to the last few millennia of
glacier cover, most of the older evidence having been overridden by ice retreat
retreat during the deglaciation\footnote{\textbf{Johan,} can we refer to one
of your papers here?}.

@@

The possibly of a widespread late-glacial in
the North American Cordilleran has been debated for long. On the southern tip
of the Coast Mountains, radiocarbon-dated frontal moraines indicate multiple
stages of glacial re-advance in the Fraser and Squamish valleys, of which
one corresponds to the Younger Dryas chronozone \citep{Clague.etal.1997,
Friele.Clague.2002, Friele.Clague.2002a, Kovanen.2002,
Kovanen.Easterbrook.2002}. In the Finlay River area, northern Rocky Mountains,
sharp-crested moraines show an abundant remobilization of late-glacial
sediments by alpine glaciers in interaction with the main body of the decaying
ice sheet \citep{Lakeman.etal.2008}. Additional clues for a late-glacial
re-advance include alpine glacier moraines in the eastern Coast Mountains,
Rocky Mountains and Columbia Mountains \citep{Menounos.etal.2008}. Although
further work is needed to constrain the timing of the late-glacial re-advance,
to seize its extent and to identify potential climatic triggers
\citep{Menounos.etal.2008}, it is interesting to note that the simulation
driven by the GRIP record produces a late-glacial re-advance in the Coast
Mountains, Rocky Mountains and Finlay River area, corresponding to where it
has been identified in the geologic record (Fig.~\ref{fig:deglac}, left panel),
whereas the EPICA-driven simulation produces a nearly-continuous deglaciation
with very restrained glacial re-advance (Fig.~\ref{fig:deglac}, right panel).

Although the predicted timing of glacier retreat differs between the two
simulations, its patterns are relatively consistent (Fig.~\ref{fig:deglac}). In
both runs, the southern half of the modelling domain, including the Puget
Lowland, the Coast and Rocky Mountains and the Interior Plateau of British
Columbia, are completely deglaciated by 12\,kyr, whereas significant ice cover
remains on the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias Mountains
in the northern half of the domain. After 12\,kyr, deglaciation proceeds
through an opening of the Liard lowland with radial margin retreat towards the
surrounding mountain ranges and inward flow towards the depression, in
compatibility with the regional melt water record \citep{Margold.etal.2013}.
Remaining ice continues its decay by retreating towards the Selwyn Mountains
and, then, the Skeena Mountains. The last remains of the palaeo-ice sheet
vanish in the Skeena Mountains at around 10.6\,kyr in the EPICA simulation
versus 9.4\,kyr in the GRIP simulation.

Once more, snapshots of modelled deglacial ice sheet configurations
(Figs.~\ref{fig:deglacshots-grip} and~\ref{fig:deglacshots-epica}%
\footnote{I produced one figure for each run, but as you could expect from
    Fig.~\ref{fig:deglac}, Figs.~\ref{fig:deglacshots-grip}
    and~\ref{fig:deglacshots-epica} look very similar. One of them would
    probably be enough, or?}) provide a
poor medium for direct comparison with geomorphological evidence of the ice
retreat. Here, we compute a map of deglacial, basal flow directions immediately
preceding exposure of ice-free land or cessation of sliding upon the approach
of the retreating margin (Fig.~\ref{fig:lastflow}). On the northern and
southern front of the ice sheet and in the Liard Lowland, the resulting map
show similarities with patterns of glacial lineations
(\citealp{Prest.etal.1968}; \citealp[Fig.~1.12]{Clague.1989};
\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~2]{Margold.etal.2013}).
However, such is not the case on the Interior Plateau of British Columbia,
where both simulations predicts neglictible basal sliding during deglaciation
(Fig.~\ref{fig:lastflow}), whereas glacial lineation maps indicate substantial
eastward flow.

On the one hand, this could indicate that the model performs poorly in this
region, perhaps because the modelled last glacial maximum ice sheet is too
thick, or too extensive to the west. Because feedback mechanisms between ice
sheet topography and regional climate such as wind redistribution, orographic
precipitation and latent warming of moisture-depleted air are absent in our
model, we regard this possibility as fully probable \citep{Seguinot.etal.2013}.
On the other hand, in both simulations driven by the GRIP and EPICA records,
the deglaciation of central British Columbia consists of a rapid northwards
retreat associated with southwards-flowing, non-sliding ice lobes on the
Interior Plateau, while the surrounding mountain ranges are already ice-free
(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). This result
appears compatible with the prevailing conceptual model of deglaciation of
central British Columbia, in which mountain ranges emerge from the ice before
the plateau \citep[Fig.~7]{Fulton.1991}. If it is valid, it may suggest that
glacial lineations on the Interior Plateau of British Columbia may be of older
age than the last glacial maximum, and may have remained intact throughout the
deglaciation.
